---
title: "Predictive Modeling for Forecasting the 2024 US Presidential Election"
subtitle: "Trump's Narrow Victory Over Harris by Less Than One Percent of the Supporting Rate"
author: 
  - Bo Tang
  - Mingjing Zhan
  - Yiyi Feng
thanks: "Code and data are available at: https://github.com/kqlqkqlqF/Investigate-Price-Freeze-in-Canadian-Grocery-Sector.git."
date: today
date-format: long
abstract: "This study presents a predictive model for the 2024 U.S. Presidential Election, focusing on the race between Donald Trump and Kamala Harris. Our model forecasts a narrow victory for Trump, estimating his average support at 44.51% compared to Harris's 43.86%, with leads of Trump in swing states. The analysis shows that state and recency are important for understanding voter support trends, reflecting the electoral system's winner-takes-all nature. This research allows electoral forecasting by demonstrating how localized support influences national outcomes and shows the need for improved polling methodologies."
format:
  pdf:
    toc: true
    number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

#### Preamble ####
# Purpose: Help constructing the "Predictive Modeling for Forecasting the 2024 US Presidential Election" paper

# Author: Bo Tang, Yiyi Feng, Mingjing Zhan
# Date: 14 November 2024
# Contact: qinghe.tang@mail.utoronto.ca, yiyi.feng@mail.utoronto.ca, mingjin.zhan@mail.utoronto.ca

####Workspace setup ####

library(dplyr)
library(here)
library(modelsummary)
library(tidyverse)
library(sf)
library(arrow)
library(ggplot2)
library(janitor)
library(purrr)
library(knitr)
library(kableExtra)
library(usmap)
library(broom)
library(caret)
library(randomForest)
library(rstanarm)

cleaned_data <- read_parquet(here::here("data/02-analysis_data/cleaned_data.parquet"))
analysis_data <- read_parquet(here::here("data/02-analysis_data/analysis_data.parquet"))

```


# Introduction

The upcoming U.S. Presidential Election marks an important point in the nation’s political landscape, shaped by public opinion, social and economic factors, and the complexities of the electoral process. With Kamala Harris and Donald Trump competing for the presidency, accurately predicting the outcome is increasingly important. Polls not only reflect voter opinion but also influence campaign strategies and media coverage. However, challenges like sampling biases, inconsistent methods, and the gap between the popular vote and the electoral vote emphasize the need for an improved forecasting model. This paper aims to develop a predictive framework that uses national polling data and examines state-level dynamics, especially in key swing states that often decide election results.

Our main focus is the probability of Donald Trump winning the 2024 U.S. presidential election, represented by voter support rates. To estimate support for both Trump and Harris, we developed linear models that account for factors such as candidate identity, poll recency, state, sample size, poll score, and poll quality, along with interactions among these variables. By identifying the optimal model, we aim to determine how these factors and their combinations influence expected support, providing insights into each candidate's chances across different regions and polling conditions. This approach models support rates rather than direct winning probabilities, allowing for a nuanced prediction that reflects variations by candidate, state, and recency.

Our model predicts that Donald Trump will win by a narrow margin, with an average support of 44.51% compared to Kamala Harris's 43.86%. Trump leads in six out of seven key swing states, suggesting that this localized support could enhance his overall chances despite only a slight national lead. Among the predictor variables analyzed, state and recency are the most significant indicators of support trends, reflecting the "winner-takes-all" nature of the U.S. electoral system and the increasing accuracy of polling data as Election Day approaches.

The remainder of this paper is structured as follows: [@sec-data] provides an overview of the dataset, details of the parameters, outcome and predictor variables, and the packages used during processing. [@sec-model] explains the modeling approach, and best model selection, justifying the choice of predictors and outlining the methods used to forecast support for Trump and Harris. [@sec-results] presents the findings, including a summary of the predicted support rate for Trump, a comparison of the predicted support rates for Trump and Harris, and a breakdown of their support rates in each state. In [@sec-discussion], we discuss the implications of these results, the limitations of our analysis, and potential avenues for future research. Additional methodological details and diagnostics are included in the appendix.



# Data {#sec-data}

## Overview

In this analysis, we used R [@citeR] to investigate polling data on public sentiment leading up to the election. Our dataset, sourced from FiveThirtyEight [@fivethirtyeight2024], provides a detailed snapshot of shifting public opinion over time. We examined key factors influencing support percentages, including poll timing, pollster characteristics, and state-specific trends.

Several R packages were vital for our data manipulation, modeling, and visualization efforts. The dplyr package provided efficient tools for data transformation and summarization [@dplyr], while modelsummary enhanced the presentation of model outputs in a clear and organized manner [@modelsummary]. We used sf for handling spatial data, enabling analysis of state-level dynamics in the election [@sf]. purrr streamlined functional programming, allowing for the application of functions across data structures [@purrr]. kableExtra created customizable tables to improve our data presentation [@kableExtra], and usmap facilitated mapping of electoral data across states [@usmap]. The broom package converted complex model outputs into tidy data frames for easier analysis [@broom]. Package caret provided a unified framework for building and evaluating machine learning models [@caret], while the randomForest package enabled the use of random forest modeling techniques for our predictive analysis [@randomForest]. Finally, testthat ensured the reliability of our analyses through code testing [@testthat]. Our workflow closely adhered to best practices, as outlined in [@tellingstories], enhancing the robustness of our predictive framework.

Our group focused on Trump’s approval ratings, aiming to ensure the credibility of the data. To achieve this, we selected only pollsters with numeric grades above 2.0, and used data collected from November 15, 2022, to October 27, 2024.

## Measurement
	



# Result {#sec-results}

```{r}
#| label: tbl-one
#| tbl-cap: "Summary Statistics of Predicted Support for Donald Trump"
#| message: false
#| echo: false
#| warning: false



```

[@tbl-one] shows the 

```{r}
#| label: tbl-two
#| tbl-cap: "Predicted Average Supporting Percentages for Donald Trump vs. Kamala Harris"
#| warning: false
#| echo: false
#| eval: true



```

[@tbl-two] presents the 

```{r}
#| label: fig-one
#| fig-cap: "Map of Projected Winner by State"
#| warning: false
#| echo: false
#| eval: true


```

[@fig-one] shows 


```{r}
#| label: fig-two
#| fig-cap: "Map of Projected Winner by State"
#| warning: false
#| echo: false
#| eval: true


```

[@fig-two] shows 





# Discussion {#sec-discussion}

## Correlation vs. Causation



## Missing Data



## Sources of Bias




\newpage

# Appendix a. {-}

## Overview of Emerson College Polling Methodology (October 23-25, 2024)

The Emerson College Polling conducted a survey from October 23 to 25, 2024, targeting 1,000 likely voters to investigate the differences in support for various candidates. In this presidential election, 58% support former President Donald Trump, while 39% support Vice President Kamala Harris.

## Population, Frame, and Sample

In this context, the target population consists of likely voters in the U.S. elections, defined by their likelihood to vote in the upcoming elections and their voting history, both of which are self-reported in the survey. The sampling frame specifically focuses on likely voters in Montana, who were reached through a combination of cell phone contacts provided by Aristotle and an online voter panel from CINT. The sample consists of 1,000 likely voters randomly selected from the sampling frame, with their status determined by a combination of voter history, registration status, and demographic data, all of which are self-reported. This methodology provides a balanced overview of Montana voters' priorities, with a credibility interval of +/- 3%.

## Sampling Approach and Trade-offs

Emerson College utilized a mixed-mode sampling approach for its poll of likely voters in Montana. This strategy involved two main methods: sending MMS text messages linked to an online survey using Aristotle’s voter lists and accessing a pre-screened, opt-in online panel from CINT. The MMS method is efficient and cost-effective, allowing participants to complete the survey at their convenience, which can enhance response rates. The online panel broadens coverage to include voters not reachable through text, capturing a wider demographic range across the state. Together, these methods create a diverse sample while reducing costs compared to traditional phone or in-person interviews.

However, this approach has trade-offs. The MMS survey requires recipients to have active cell phones and internet, potentially excluding older or less tech-savvy voters. Additionally, the online panel consists of self-selected participants, which may not fully reflect the general voter population. Mixing data from both sources can introduce inconsistencies, as each method may attract different respondent types, necessitating careful weighting to maintain balance and accuracy. Smaller demographic subsets, such as age, race, or education, carry higher credibility intervals due to reduced sample sizes, limiting precision in analysis. Overall, the mixed approach optimizes reach, reduces costs, and shows the priority needs of Montana’s voters, although there are limitations.

## Non-response Handling

Emerson College does not provide specific details regarding its non-response management. While it mentions that data were weighted by demographics such as gender, education, race, age, party registration, and region to align with the 2024 likely voter model, this weighting primarily addresses demographic imbalances and does not directly mitigate non-response bias. The survey lacks information on common non-response strategies, such as follow-up attempts, participation incentives, or specific adjustments for non-responders. This absence raises concerns about potential non-response bias, particularly if certain demographic groups were less likely to engage with the survey.

## Questionnaire Design

This questionnaire has strong points. Its straightforward and clear wording makes questions easy for respondents to follow and reduces potential confusion. By focusing on issues like the economy, housing, and voter approval for specific candidates, it captures key voter concerns in Montana, offering a concise view. The use of multiple questions around candidate approval, voter issues, and demographics adds depth to the questionnaire.
However, the questionnaire also has limitations. While demographic questions enhance the survey’s representativeness, smaller groups (e.g., nonbinary individuals) may carry higher credibility intervals, reducing precision for those subgroups. The mixed-mode approach (online panel and mobile) improves access but still risks non-response bias, as certain demographics might be less likely to participate. Overall, the design achieves clarity and breadth, though response biases and sample variations should be considered in interpreting the findings. For example, in this survey of 1,000 Montana voters, only 5 respondents identified as nonbinary or other genders. Since statistical reliability depends on the number of responses, small groups have higher variability, meaning their responses can swing widely due to each individual answer carrying greater weight.

# Appendix b. {-}

## Idealized Methodology for Forecasting the U.S. Presidential Election

We aim to develop a methodology for forecasting U.S. presidential election outcomes by conducting a survey with a $100,000 budget. Using stratified random sampling and multi-mode recruitment, the survey targets 10,000 likely voters across demographic and regional lines. Key measures include data validation checks, weighted analysis, and predictive modeling to ensure accuracy. Results will be enriched by aggregating reputable data sources like FiveThirtyEight for a better forecast.

## Budget Allocation

Funding allocations will focus on ensuring thorough and effective sampling, recruitment, data validation, and analysis methods are employed with a total budget of no more than $10,000. The proposed budget breakdown is as follows:

Survey platform costs: $10,000 (subscription fees for online survey tools such as Google Forms or Qualtrics)

Respondent incentives: $10,000 (gift cards or other incentives to encourage participation)

Recruitment and staffing: $35,000 (staffing costs for survey distribution and data collection)

Data analysis tools: $20,000 (statistical software licenses, data cleaning and analysis)

Marketing and promotion: $10,000 (awareness and engagement campaigns)

Contingency fund: $5,000 (for unexpected expenses)

## Sampling Approach

The sampling approach will employ a stratified random sampling method to ensure representation across various demographic groups, including age, gender, race, education level, geographical location, and party affiliation. The target population consists of likely voters in the U.S., defined by historical voting behavior and self-reported intentions to vote. A sample size of approximately 10,000 respondents will be aimed at ensuring statistical robustness and a credibility interval of +/- 1% at a 95% confidence level.
This will be achieved through a combination of national voter registration databases to identify potential respondents. For example, we can utilize the National Voter Registration Act (NVRA) data from the National Association of Secretaries of State (NASS) to access information on registered voters. This database will allow us to filter for likely voters based on their registration status and historical voting behavior, ensuring that our sampling frame is representative of the electorate. By using reliable sources, we can create a sampling framework that enhances the accuracy of our election forecasts.

## Respondent Recruitment and Data Validation

To reach the target population, a multi-mode recruitment strategy will be implemented:
Online Surveys: Use Google Forms to distribute the survey electronically.
Telephone Surveys: Conduct live telephone interviews to capture demographics that might not engage online.
Text Messaging Surveys: Implement SMS surveys to reach younger demographics and those without regular internet access.
Incentives: Offer gift cards or other small incentives for participation, particularly for online respondents. This can enhance response rates and engagement.

To effectively reach our target population and capture a diverse range of perspectives, we will implement a multi-mode recruitment strategy tailored to different demographic groups and communication preferences. This approach includes online surveys using a Google Forms questionnaire, which will be widely distributed through social media, email lists, and community networks to maximize reach among individuals who frequently engage online. The user-friendly platform allows participants to complete the survey quickly and anonymously on any internet-enabled device. The survey can be accessed through the following link: [https://forms.gle/oSbad52Vuw9Z9Wf46](https://forms.gle/oSbad52Vuw9Z9Wf46). Additionally, we will conduct live telephone interviews to include participants who may not be reachable through online channels, ensuring we capture responses from populations that might otherwise be underrepresented. To further engage younger demographics and individuals with limited internet access, we will implement SMS-based surveys, allowing participants to respond quickly via text. To encourage participation and improve response rates, small incentives such as digital gift cards will be offered, particularly for online respondents, with details communicated at the survey's start and awarded upon completion to ensure transparency. This comprehensive approach will enable us to gather a robust and representative dataset, providing valuable insights into the preferences and priorities of voters across multiple demographics.

## Poll Aggregation and Modeling

Poll results will be aggregated using statistical methods to identify trends and analyze historical voting patterns. For model building, we will implement a weighted analysis to ensure demographic representation, applying specific weights based on factors such as age, gender, race, and education level, as well as state significance to reflect regional variations in voter behavior. Predictive analytics will primarily involve logistic regression to model voting preferences and forecast election outcomes, supplemented by time-series analysis to track changes in voter sentiment over the campaign period. To enhance our findings, we will combine our data with reputable sources like FiveThirtyEight, utilizing their aggregation techniques to enrich our analysis. 

\newpage

# References

